To be FAIR: Theory Specification Needs an Update

### 2025-06-08 action_letter.Rmd
update 2025-06-08 13:02:50.777903

---
title             : "To be FAIR: Theory Specification Needs an Update"
authors           : "Caspar van Lissa on behalf of co-authors"
journal           : "Perspectives on Psychological Science"
handling_editor   : "Vrieze"
class             : "draft"
output            : rmarkdown::html_document
---

<!-- revise::revise_letter_pdf -->

@@

```{r setup-chunk, include = FALSE}
library(revise)
revise::read_manuscript("abstract.md")
revise::read_manuscript("manuscript.Rmd")
```

@@

<!-- AB: -->
1.	I agree that striking the balance between the conceptual and technical details is the very challenge of this paper. I think we should go a route in the revision in which we try to explain technical details in even simpler language. For example, the github reference to our code could read like this: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX." And maybe there is a way to move the more technical details to a section that can be skipped without compromising the conceptual/opinion aspects.

@@

We thank the Reviewer for their thoughtful review and detailed comments. We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.

@@

Thank you for this suggestion; as proposed, we now briefly mention the R-package in the abstract:

@@

nakagawaPoorHypothesesResearch2024

@@

<!-- AB: -->
4.	What is a theory? (Suggestion to move up the section) It seems to always come back to this that readers a confused because everyone has their own idea of what a theory is. We need to think of one introductory sentence very early on that basically sets the stage and tells people no matter what they think a theory is they very likely will find the FAIRification useful 🙂

@@

To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
We also already have two worked examples for actual theories, which we now point out in the text (with links to the relevant vignettes).

@@

We appreciate this thoughtful and provocative comment, which explores the broader sociological and cultural dimensions of scientific reform.
We are sympathetic to the reviewer’s view that the challenges facing psychology are not solely technical or procedural, but likely also reflect deeper issues of norms and values.

That said, we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
As methodologists, our primary aim in this paper is to articulate what FAIR theory affords: a framework that supports better theory specification, sharing, and reuse.
We believe this contribution is most effective when kept sharply focused on the practical infrastructure and its scientific use cases.

@@

We are sympathetic to the Reviewer's meta-scientific considerations,
and we hope that the Reviewer will agree that - while there is clearly a need for enquiry into norms and values - the present paper might not be the best avenue for it.

>>>>

---
title             : "To be FAIR: Theory Specification Needs an Update"
authors           : "Caspar van Lissa on behalf of co-authors"
journal           : "Perspectives on Psychological Science"
handling_editor   : "Vrieze"
class             : "draft"
bibliography      : "fair_theory.bib"
output            : revise::revise_letter_pdf
---

@@

```{r setup-chunk, include = FALSE}
library(revise)
rm(".revise_manuscripts")
revise::read_manuscript("abstract.md")
revise::read_manuscript("test.Rmd")
```

@@

<!-- AB: -->
1.	I agree that striking the balance between the conceptual and technical details is the very challenge of this paper. I think we should go a route in the revision in which we try to explain technical details in even simpler language. For example, the github reference to our code could read like this: "Our manuscript is reproducible given the code provided in our Github repository identified by the snapshot with the unique commit id #XXXXX." And maybe there is a way to move the more technical details to a section that can be skipped without compromising the conceptual/opinion aspects. @wilkinsonFAIRGuidingPrinciples2016

@@

We thank the Reviewer for their thoughtful and encouraging review.
We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory—not only within psychology, but also in related disciplines.
We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.

@@

As suggested, we now briefly mention the R-package in the abstract:

@@

We thank the Reviewer for pointing this out; we have included the reference to the paper making this argument in the following section:

`r get_revision("waste")`

@@

<!-- AB: -->
4.	What is a theory? (Suggestion to move up the section) It seems to always come back to this that readers a confused because everyone has their own idea of what a theory is. We need to think of one introductory sentence very early on that basically sets the stage and tells people no matter what they think a theory is they very likely will find the FAIRification useful

@@

To address this comment, we have removed as much of the specific detail as possible from the main text, and instead refer to the package vignettes (we have created an additional vignette to make this change possible, and extended several other vignettes).
There are already two worked examples for actual theories in the package vignettes,
which we now reference in the text.

The resulting shortened section reads as follows:

`r get_revision("tutorial")`

@@

We appreciate this thought-provoking comment, which explores the broader sociological and cultural dimensions of scientific reform.
We are sympathetic to the reviewer’s view that the challenges facing psychology are not solely technical or procedural, but likely also reflect deeper issues of norms and values.

That said, we are not sure that we are best positioned to engage with these issues, nor that the present paper would be the best place for such an argument.
As methodologists, our primary aim in this paper is to articulate the affordances of FAIR theory:
a framework that supports better theory specification, sharing, and reuse.
We believe our contribution will be most effective if it remains sharply focused on the practical infrastructure and its scientific use cases.

@@

We share the Reviewer's interest in the meta-scientific perspective,
and we hope that the Reviewer will agree that - while there is clearly a need for enquiry into norms and values - the present paper might not be the best avenue for it.

We did make one minor change to address this comment, and "punched up" the section on Theory and Scientific Progress, with a reference to scientific norms as documented in ethical guidelines:

`r get_revision("norms")`

### 2025-06-08 test.Rmd
update 2025-06-08 13:02:50.777903

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
<!-- CJ: I agree, but the point about phenomena detection is not directly relevant to this paper, so I prefer not to get into it here. -->
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978, p. 1].
The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

We propose a practical, software-agnostic workflow for making theory FAIR.
While these steps can be implemented using a variety of tools,
R users can streamline the process using the `theorytools` package, which automates most steps.
We will illustrate the workflow using De Groot's *empirical cycle*,
a meta-theory of theory construction,
as a running example for this tutorial.
The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.

@@

* Set the *resource type* to the category `Model`,
* Add the words `FAIR theory:` to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses are encouraged to use the words `meta-analysis` in the title),
* Add `fairtheory` to the keywords to aid search engine indexation.
* Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.

>>>>

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
There are, however, indications that contemporary psychology falls short of this ideal.
Meehl observed that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978, p. 1].
Recent empirical findings confirm this view.
Firstly, because hypothesis-testing research is vastly over-represented in the literature, amounting to 89.6% of published papers [@kuhbergerPublicationBiasPsychology2014].
Closer examination of such studies reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of hypothesis-testing studies referenced any theory,
and rarely in direct relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to theory.

Perhaps some ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are not of substantive interest, but merely reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
such as straw-man null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
<span id = "waste">Testing ad-hoc hypotheses not grounded in theory, or grounded in misinterpreted- or multi-interpretable theory,
does not advance our principled understanding of psychological phenomena,
and consequently contributes to research waste [@nakagawaPoorHypothesesResearch2024].</span>
<span id = "norms">Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021]; inefficient at best, and more likely, futile.
As the Declaration of Helsinki prescribes that ethical (medical) research with human participants must "avoid research waste",
our field should take seriously its ethical responsibility to develop procedures to reduce it.</span>
The present paper does so by introducing procedures to improve transparent and unambiguous communication about theory; 
instantiating theory as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

<span id="tutorial">We propose a practical, software-agnostic workflow for making theory FAIR.
While these steps can be implemented using a variety of tools,
R users can streamline the process using the `theorytools` package, which automates most steps.
We will illustrate the workflow using De Groot's *empirical cycle*,
a meta-theory of theory construction,
as a running example for this tutorial.
The resulting FAIR implementation of this theory is available at <https://doi.org/10.5281/zenodo.14552329>.

@@

* Set the *resource type* to the category `Model`,
* Add the words `FAIR theory:` to the title so that sentient readers will recognize the work as a FAIR theory (just as meta-analyses are encouraged to use the words `meta-analysis` in the title),
* Add `fairtheory` to the keywords to aid search engine indexation.
* Optionally, submit the theory to the ["FAIR Theory Community"](https://zenodo.org/communities/fairtheory) to contribute to community building; communities on Zenodo are shared spaces to manage and curate research outputs.
</span>
